 

### 2021-06-26 vertical-thermal-review.Rmd
worked on leafT section revamp

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).

@@

**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, but during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits (see next section), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).
*(I want to add a few sentences on how this varies with environmental conditions-- drought, high humidity....)*

(**Vertical gradients in $T_{leaf}-T_{air}$ are expected to vary with midrometeorlogical conditions.**)

~~~ (*I think we should cut everything between here and  next "~~~", after harvesting the most important content for the section above.*)

Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. In contrast, when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to greater $g_a$ than larger leaves that take longer due to lower $ga$. Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

~~~

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1]. However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020]. Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]

>>>>

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed [Fig. `r fig_leaf_T`b,@daudetWindSpeedLeaf1999a] and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with with vapor pressure deficit, or increases with RH [Fig. `r fig_leaf_T`c].

@@

**Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].** 
When $g_s$ is high, high $\lambda E$ facilitates heat dissipation. 
However, when inadequate water supply -- for example under drought -- causes leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [Fig. `r fig_leaf_T`e, @fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads, but during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
With adequate water, species with smaller leaves (higher $g_a$) and larger $g_s$ will be most effective at cooling during high temperatures [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits (see next section), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).

**Vertical gradients in the propensity for $T_{leaf}$ to be elevated above $T_{air}$ are expected to vary with micrometeorlogical conditions (Fig. `r fig_leaf_T`).**
Under hot conditions, higher wind speeds and $g_a$ enable cooling, but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf} - T_{air}$ [Fig. `r fig_leaf_T`b, @martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]. 
Thus, understory leaves, which typically experience lower windspeeds (Fig. `r fig_NEON_vertical`), would retain more heat.
The higher-RH conditions of the understory (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
Thus, understory leaves would have less ability to shed excess heat under hot and humid conditions.
In contrast, under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a propensity for $T_{leaf} - T_{air}$ to be elevated more in the canopy than the understory (Fig.`r fig_leaf_T`).

~~~ **(KAT stopped here)**

 Across species, inherent leaf $gs$ is generally greater in sun leaves than shade leaves, due to stomata density [Table 1]. However, because factors driving $g_s$ limitation--irradiance, $VPD$ and $T_{air}$--including increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, resulting in increased in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ and thermal capacitance can result in greater $T_{leaf}$ under slight heat stress [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020]. Lastly, canopy structure and architecture adaptation can also influence canopy temperature  [@leuzingerTreeSpeciesDiversity2007; @mullerEvidenceEfficientNonevaporative2021]
